A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-07-11 tidyindex.qmd
a new draft version of the introduction

**Why index is useful, why people care about indexes**

*incorporate the following in why using index: multiple pieces of information (variables) that need to be taken into account*

Many concepts relevant to decision making cannot be directly measured, however, they are crucial for resource allocation, early prevention, and other operational purpose.

  - USDM and European Drought Observatory publishes drought indicator to [...], 
  - NOAA and BOM producing forecast on El Nino alert system to [...]
  - OECD and UN producing indexes measuring socio-economic aspects of country for policy making.

There decision making often needs to consider multiple aspects and indexes is a statistical tool to pack these information into a single number, making it easier for policy makers and public communication.

While the [decision to make] is often multi-faceted, simple indexes tend to be favoured when it comes to communicating to stakeholders and public.

**Define what is an index, what is not**

While indexes are commonly used in operation and research, their compositions and structures can vary substantially. An index can be composed from a singe variable, such as the Standardized Precipitation Index (SPI) [@spi], or multiple variables, also known as composite indexes or composite indicators, like the Human Development Index [@hdi]. Some indexes can be expressed as a single equation, like Normalized Difference Vegetation Index (NDVI) [@ndvi], calculated as $(\text{NIR} + \text{Red})/(\text{NIR} + \text{Red})$. In other cases, indexes themselves can be used as variables to construct indexes. For example, SPI is used to in many complex indexes, such as the U.S. Drought monitor and European Combined Drought Index (CDI) [@cdi]. While the index framework proposed in this paper applies to all the aforementioned indexes, our focus will be on studying complex indexes that combine multiple variables using statistical methods.

The initial plan aims to investigate how variable combinations can affect the indexes and whether we can use statistical methods to recommend better combination for decision making. However, when analysing specific index examples, we find indexes are often not just a simple weighted linear aggregation of variables. When indexes have a spatial and/or temporal component, different choices on aggregating space and time will introduce uncertainty. [There is always new variables found to be related to make the decision and should be introduced into the index. The decision made in pre-processing data coming from different sources such as satellite imageries, in-situ stations, and large physical models.] While numerous indexes are proposed in the literature, the well-known ones tend to prefer a simple procedure: they may only consider one or two variables, i.e. SPI or use a decision-tree based process to incorporate multiple variables, i.e. (How El Nino is monitored by major climate institutes). More complex indexes often have difficulties to reach consensus, due to their uncertainty when applied in different scenarios.

**What is the challenges with current index construction**/ **what can be done if people adopt this pipeline/ why it is beneficial?**/**who would benefit from this paper**

Typical ways to evaluate indexes is to compare it with the standard/ well-known indexes, such as SPI or SPEI for drought indexes, or a response variable relevant for the monitoring purpose, i.e. crop yield for monitoring agricultural drought. While these evaluations show the accuracy of the index to capture, they doesn't address the robustness of the index,[ given variation of the input parameters, or a change of different study areas or across time.] To evaluate the robustness of indexes, indexes need to be broken down into their fundamental building blocks, and then analyse the effect of each component. The OECD handbook [@oecd_handbook_2008] has recommended a set of guidelines for construct composite indexes for measuring country's socio-economic performance. This paper identifies a more general pipeline framework from which indexes can be built from. Based on the pipeline, an R package, tidyindex, is developed to [validate the indexes]. This work provides researchers actively developing new indexes with tools to evaluate the proposed indexes. It also helps index analysts to identify weakness in existing indexes for methodology improvement.

@@

@@

>>>>

  <!-- - Why index is useful, why people care about indexes -->
  <!-- - Define what is an index, what is not -->
  <!-- - What is the challenges with current index construction -->
  <!-- - what can be done if people adopt this pipeline/ why it is beneficial? -->
  <!-- - who would benefit from this paper -->

Indexes are commonly-used tools to monitor, communicate and make decisions across various domains in society. Their applications can be found in the environment (e.g. Air Quality Index, El Niño-Southern Oscillation Index), in the economy (Consumer Price Index, stock market indexes), and in the social sciences (e.g. QS University Rankings, Human Development Index).

To construct an index, experts typically first define an underlying concept of interest to measure, then collect relevant variables for measuring the concept, and lastly summarise these variables into a single numerical or categorical value. An index can be as simple as counting the number of days when the maximum temperature is greater than 25 degree. This is "the number of summer days" (SU), a climate index listed in ETCCDI Climate Change Indices [@climate-indexes]. In some cases, indexes are weighted linear combinations of variables. For example, Human Development Index (HDI) [@hdi] combines the three aspects (life expectancy, years of schooling, and GNI per capita) to measure country's development. In more complex scenarios, indexes are constructed by combining multiple variables, often accompanied by spatial and temporal considerations arising from diverse data formats, including in-situ stations, satellite imageries, and model outputs. An example of this is the Combined Drought Index (CDI) [@cdi], curated by the European Drought Observatory, which uses a decision tree to combine Standardized Precipitation Index (SPI) [@spi] with the satellite measure, Fraction of Absorbed Photosynthetically Active Radiation (FAPAR), and modelled soil moisture.

Our initial interest is to understand how different variable combinations contribute to the uncertainty of an index. However, as we start to understand how indexes are composed, we find many other steps in the construction also introduce uncertainty. In order to evaluate the uncertainty, an index needs to be broken down into its fundamental building blocks to analyse the effect of each component. This shifts the focus to investigate the decomposition of indexes into their constituent parts for analysing uncertainty. While the OECD handbook [@oecd_handbook_2008] has provided a set of steps for constructing composite socio-economic indexes, there is a need to extend these guidelines to accommodate more complex construction steps required for indexes in general.

In this work, we develop a data pipeline framework where indexes from different domains can be built from. Based on the pipeline, an R package, tidyindex, is developed for practitioners to construct indexes and understand how an index responds to changes in different components in the pipeline. This work provides researchers actively developing new indexes with tools to evaluate the proposed indexes on their robustness for wider adoption. It also helps index analysts to diagnose an index to identify its weakness for methodology improvement.

@@

<!-- Evaluating indexes is challenging because the underlying concept an index measures is often a latent variable and lacks direct benchmarks for comparison. In literature, researchers typically demonstrate the accuracy of an index through comparing it to some well-known indexes, such as SPI for drought indexes, or to a relevant response variable, such as crop yield for monitoring agricultural drought. For social indexes, @uncertainty proposes a Monte Carlo algorithm to compute index uncertainty through sampling the linear weights.  -->

<!-- These sometimes make indexes difficult to reach consensus for wider use and large institutes often prefer simple indexes for more stable performance and easier interpretation. -->

<!-- When indexes have a spatial and/or temporal component, different choices on how space and time is aggregated will introduce uncertainty. A slight change in the variable value can sometimes introduce a big difference for the index.  -->

@@

Future work:
  - integrate more complex dimension reduction methods to calculate weights 
  - strengthen the spatial processing module

### 2023-07-11 tidyindex.tex
a new draft version of the introduction

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, interior hidden, boxrule=0pt, enhanced, breakable]}{\end{tcolorbox}}\fi

@@

\textbf{Why index is useful, why people care about indexes}

\emph{incorporate the following in why using index: multiple pieces of
information (variables) that need to be taken into account}

Many concepts relevant to decision making cannot be directly measured,
however, they are crucial for resource allocation, early prevention, and
other operational purpose.

\begin{itemize}
\tightlist
\item
  USDM and European Drought Observatory publishes drought indicator to
  {[}\ldots{]},
\item
  NOAA and BOM producing forecast on El Nino alert system to
  {[}\ldots{]}
\item
  OECD and UN producing indexes measuring socio-economic aspects of
  country for policy making.
\end{itemize}

There decision making often needs to consider multiple aspects and
indexes is a statistical tool to pack these information into a single
number, making it easier for policy makers and public communication.

While the {[}decision to make{]} is often multi-faceted, simple indexes
tend to be favoured when it comes to communicating to stakeholders and
public.

\textbf{Define what is an index, what is not}

While indexes are commonly used in operation and research, their
compositions and structures can vary substantially. An index can be
composed from a singe variable, such as the Standardized Precipitation
Index (SPI) (McKee, Doesken, and Kleist 1993), or multiple variables,
also known as composite indexes or composite indicators, like the Human
Development Index (United Nations Development Programme 2022). Some
indexes can be expressed as a single equation, like Normalized
Difference Vegetation Index (NDVI) (Tucker 1979), calculated as
\((\text{NIR} + \text{Red})/(\text{NIR} + \text{Red})\). In other cases,
indexes themselves can be used as variables to construct indexes. For
example, SPI is used to in many complex indexes, such as the U.S.
Drought monitor and European Combined Drought Index (CDI)
(Sepulcre-Canto et al. 2012). While the index framework proposed in this
paper applies to all the aforementioned indexes, our focus will be on
studying complex indexes that combine multiple variables using
statistical methods.

The initial plan aims to investigate how variable combinations can
affect the indexes and whether we can use statistical methods to
recommend better combination for decision making. However, when
analysing specific index examples, we find indexes are often not just a
simple weighted linear aggregation of variables. When indexes have a
spatial and/or temporal component, different choices on aggregating
space and time will introduce uncertainty. {[}There is always new
variables found to be related to make the decision and should be
introduced into the index. The decision made in pre-processing data
coming from different sources such as satellite imageries, in-situ
stations, and large physical models.{]} While numerous indexes are
proposed in the literature, the well-known ones tend to prefer a simple
procedure: they may only consider one or two variables, i.e.~SPI or use
a decision-tree based process to incorporate multiple variables,
i.e.~(How El Nino is monitored by major climate institutes). More
complex indexes often have difficulties to reach consensus, due to their
uncertainty when applied in different scenarios.

\textbf{What is the challenges with current index construction}/
\textbf{what can be done if people adopt this pipeline/ why it is
beneficial?}/\textbf{who would benefit from this paper}

Typical ways to evaluate indexes is to compare it with the standard/
well-known indexes, such as SPI or SPEI for drought indexes, or a
response variable relevant for the monitoring purpose, i.e.~crop yield
for monitoring agricultural drought. While these evaluations show the
accuracy of the index to capture, they doesn't address the robustness of
the index,{[} given variation of the input parameters, or a change of
different study areas or across time.{]} To evaluate the robustness of
indexes, indexes need to be broken down into their fundamental building
blocks, and then analyse the effect of each component. The OECD handbook
(OECD, European Union, and Joint Research Centre - European Commission
2008) has recommended a set of guidelines for construct composite
indexes for measuring country's socio-economic performance. This paper
identifies a more general pipeline framework from which indexes can be
built from. Based on the pipeline, an R package, tidyindex, is developed
to {[}validate the indexes{]}. This work provides researchers actively
developing new indexes with tools to evaluate the proposed indexes. It
also helps index analysts to identify weakness in existing indexes for
methodology improvement.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ndvi}{}}%
Tucker, Compton J. 1979. {``Red and Photographic Infrared Linear
Combinations for Monitoring Vegetation.''} \emph{Remote Sensing of
Environment} 8 (2): 127--50.
\url{https://doi.org/10.1016/0034-4257(79)90013-0}.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, sharp corners, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

Indexes are commonly-used tools to monitor, communicate and make
decisions across various domains in society. Their applications can be
found in the environment (e.g.~Air Quality Index, El Niño-Southern
Oscillation Index), in the economy (Consumer Price Index, stock market
indexes), and in the social sciences (e.g.~QS University Rankings, Human
Development Index).

To construct an index, experts typically first define an underlying
concept of interest to measure, then collect relevant variables for
measuring the concept, and lastly summarise these variables into a
single numerical or categorical value. An index can be as simple as
counting the number of days when the maximum temperature is greater than
25 degree. This is ``the number of summer days'' (SU), a climate index
listed in ETCCDI Climate Change Indices (Zhang 2020). In some cases,
indexes are weighted linear combinations of variables. For example,
Human Development Index (HDI) (United Nations Development Programme
2022) combines the three aspects (life expectancy, years of schooling,
and GNI per capita) to measure country's development. In more complex
scenarios, indexes are constructed by combining multiple variables,
often accompanied by spatial and temporal considerations arising from
diverse data formats, including in-situ stations, satellite imageries,
and model outputs. An example of this is the Combined Drought Index
(CDI) (Sepulcre-Canto et al. 2012), curated by the European Drought
Observatory, which uses a decision tree to combine Standardized
Precipitation Index (SPI) (McKee, Doesken, and Kleist 1993) with the
satellite measure, Fraction of Absorbed Photosynthetically Active
Radiation (FAPAR), and modelled soil moisture.

Our initial interest is to understand how different variable
combinations contribute to the uncertainty of an index. However, as we
start to understand how indexes are composed, we find many other steps
in the construction also introduce uncertainty. In order to evaluate the
uncertainty, an index needs to be broken down into its fundamental
building blocks to analyse the effect of each component. This shifts the
focus to investigate the decomposition of indexes into their constituent
parts for analysing uncertainty. While the OECD handbook (OECD, European
Union, and Joint Research Centre - European Commission 2008) has
provided a set of steps for constructing composite socio-economic
indexes, there is a need to extend these guidelines to accommodate more
complex construction steps required for indexes in general.

In this work, we develop a data pipeline framework where indexes from
different domains can be built from. Based on the pipeline, an R
package, tidyindex, is developed for practitioners to construct indexes
and understand how an index responds to changes in different components
in the pipeline. This work provides researchers actively developing new
indexes with tools to evaluate the proposed indexes on their robustness
for wider adoption. It also helps index analysts to diagnose an index to
identify its weakness for methodology improvement.

@@
